---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# mpmtools: Tools for Building and Analyzing Matrix Population Models 

**mpmtools** provides a collection of tools for building and analyzing matrix population models
    (MPMs). An MPM represents the demography of an age-, stage-, or size-structured
    population in one or more projection matrices. This package contains tools that
    help ease the proces of constructing MPMs of varying degrees of complexity, and
    provides functions for advanced model analysis. mpmtools is intended to help 
    non-modelers more easily construct MPMs that reliably represent their data, and to
    perform analyses that would otherwise require programming skills.

## Installation

You can install mpmtools from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("BruceKendall/mpmtools")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(mpmtools)

# Create a demography schedule, with juvenile and senescent age classes
demog_sched <- data.frame(x = 0:7,
                          sx = c(0.05, 0.2, 0.35, 0.8, 0.9, 0.9, 0.75, 0.4),
                          mx = c(0, 0, 0, 0.5, 1, 3, 3, 1.5))

# Construct a Leslie matrix from this demography schedule
A1 <- make_Leslie_matrix(demog_sched)
A1

# Calculate the asymptotic growth rate of the population governed by this demography
#   schedule:
lambda1(A1)
```
